---
title: 'MIE237'
author: "Neil Montgomery"
date: "2016-02-23"
output: 
  ioslides_presentation: 
    css: 'styles.css' 
    widescreen: true 
    transition: 0.001
---
\newcommand{\Var}[1]{\text{Var}\left( #1 \right)}
\newcommand{\E}[1]{E\left( #1 \right)}
\newcommand{\Sample}[1]{#1_1,\ldots,#1_n}
\newcommand{\od}[2]{\overline #1_{#2\cdot}}
\newcommand{\flist}[2]{\{#1_1, #1_2, \ldots, #1_#2\}}
\newcommand{\samp}[2]{#1_1, #1_2, \ldots, #1_#2}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\ve}{\varepsilon}
\newcommand{\bs}[1]{\boldsymbol{#1}}



```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

# regression

## The sum of squares decomposition revisited

$$\begin{align}
SST &= \sum_{i=1}^n\left(\hat y_i - \overline y\right)^2 +
\sum_{i=1}^n\left(y_i - \hat y_i\right)^2\\
&= SSR + SSE
\end{align}$$

They are all sums of squares of normal distributions, so they have $\chi^2$ distributions with degrees of freedom: $n-1$, $1$, and $n-2$, respectively. 

In addition, $SSR$ and $SSE$ are independent, so that:
$$\frac{SSR/1}{SSE/(n-2)} \sim F_{1, n-2}$$

## Hypothesis test for the slope parameter - revisited

Main hypothesis test: $H_0:\beta_1=0$ versus $H_1:\beta_1\ne 0$. 

Key fact 1:
$$T=\frac{\hat\beta_1 - \beta_1}{\sqrt{MSE}/\sqrt{S_{xx}}}\sim t_{n-2}$$

Key fact 2:
$$F = \frac{SSR/1}{SSE/(n-2)} \sim F_{1, n-1}$$

(And actually *in the case of simple regression* $T^2 = F$ (algebraically!). The p-value will be *identical*.)

## Example - simulated from 2016-02-02

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

x <- 1:20/10
sigma <- 3
set.seed(1)
y <- 5 + 2*x + rnorm(length(x), 0, sigma)

regr_data <- data.frame(x=x, y=y)

regr_lm <- lm(y ~ x, regr_data)

summary(regr_lm)
```

## More from 2016-02-03

```{r}
anova(regr_lm)
```

## Example from 2015 exam (2016-02-05)

```{r, message=FALSE}
library(dplyr)
rust_level <- function(x) {
  rust <- c("low", "medium", "high")
  if(x < 100) {
    r <- sample(rust, 1, prob=c(0.6, 0.3, 0.1))
  } else if(x < 140) {
    r <- sample(rust, 1, prob=c(0.2, 0.6, 0.2))
  } else {
    r <- sample(rust, 1, prob=c(0.1, 0.3, 0.6))
  }
}

set.seed(1)
n <- 400
error <- rnorm(n, 0, 1)

age <- round(rnorm(n, 120, 20), 0)
max_kpa <- 200 + 50*rbeta(n, 2, 1)
min_kpa <- 400 - max_kpa + rnorm(n,0, 10)
tot_gas <- age*400 + rweibull(n, 2, 10000)
rust <- factor(sapply(age, rust_level), levels=c("low", "medium", "high"), ordered=T)
brand <- factor(sample(c("A", "B"), n, repl=T, prob=c(0.3, 0.7)))
volts <- -10 + 
  0.005 * tot_gas/age + 
  0.02 * max_kpa - 
  0.02 * min_kpa + 
  0.29 * as.numeric(rust) + 
  0.07 * as.numeric(brand) + 
  error

meters <- data.frame(age, max_kpa, min_kpa, tot_gas, volts, rust, brand)
k_lm <- lm(volts ~ max_kpa, data = meters)
summary(k_lm)
```

## More from 2015 exam

```{r}
anova(k_lm)
```

Note that the $T^2 = F$ is a bit of mathematcal trivia that applies in simple regression only. 

## New topic: $R^2$ { .build }

The "fit" of a linear model can be summarized by a single number (!):

$$\begin{align*}
SST &= SSR + SSE \\
1 &= \frac{SSR}{SST} + \frac{SSE}{SST}\\
R^2 &= \frac{SSR}{SST}\end{align*}$$

This is a moderately useful number that also goes by a unfortunately dramatic-sounding "coefficient of determination" and can be interpreted as "the proportion of variation explained by the model". 

But in the end it is just a single number that summarizes an *entire bivariate linear relationship*, so don't take it too seriously. 

## More examples

```{r, message=FALSE, fig.height = 3, fig.width = 3}
library(dplyr)
library(ggplot2)
set.seed(2)
n <- 40
x <- seq(1, 10, length.out = n)
y1 <-  1 + x + rnorm(n, 0, 0.1)
data.frame(x, y1) %>% 
  ggplot(aes(x=x, y=y1)) + geom_point() + labs(title = paste0(expression(R^2), " = ", cor(x, y1)^2))

y2 <- 1 + x + rnorm(n, 0, 1)
data.frame(x, y2) %>% 
  ggplot(aes(x=x, y=y2)) + geom_point() + labs(title = paste0(expression(R^2), " = ", cor(x, y2)^2))

y3 <- 1 + x + rnorm(n, 0, 5)
data.frame(x, y3) %>% 
  ggplot(aes(x=x, y=y3)) + geom_point() + labs(title = paste0(expression(R^2), " = ", cor(x, y3)^2))

```

## Limitations: "Model assumptions"

Assumes linear model is appropriate to begin with.

```{r, fig.height = 4, fig.width = 4}
set.seed(5)
y4 <- 1 + x + rnorm(n, 0, 2)
data.frame(x, y4) %>% 
  ggplot(aes(x=x, y=y4)) + geom_point() + labs(title = paste0(expression(R^2), " = ", cor(x, y4)^2)) 

y5 <- 20 - (x - 8)^2 + rnorm(n, 0, 8)
data.frame(x, y5) %>% 
  ggplot(aes(x=x, y=y5)) + geom_point() + labs(title = paste0(expression(R^2), " = ", cor(x, y5)^2)) 
```

## Limitations: sample size

